---
title: "Statistical Inference Course Project Part 2 - Kevin O'Leary"
geometry: tmargin=1cm, lmargin=1cm, rmargin=1cm, bmargin = 1.5cm
output:
  pdf_document:
    fig_crop: no
    fig_height: 3.55
    fig_width: 6
  html_document: default
fontsize: 5pt
---

Now we're going to analyze the ToothGrowth data in the R datasets package which 
concerns the length of odontoblasts (cells responsible for tooth growth) in 60 
guinea pigs, each receiving one of three dose levels of Vitamin C (0.5, 1.0, and
2.0 mg) with one of two delivery methods (orange juice or an aqueous solution of
ascorbic acid).  

**1. Load the ToothGrowth data and perform some basic exploratory data analyses** 

```{r}
library(datasets)
str(ToothGrowth)
```

We can see that 'dose' is a numeric vector when in fact the contents can take
only one of three levels. It is necessary then to convert this vector to a factor 
vector.

```{r}
ToothGrowth$dose <- factor(ToothGrowth$dose)
summary(ToothGrowth)
```

```{r echo=FALSE}
options("scipen"=100) ##disable scientific notation
```

Let's take a graphical look at the data:

```{r}
library(ggplot2)

ojsubset <- ToothGrowth[ToothGrowth$supp == "OJ",] ##subset data where supp = OJ
orangejplot <- ggplot(ojsubset, aes(x=dose,y=len, fill=dose)) +
    geom_boxplot() +
    theme(plot.title = element_text(size = 10),
         text = element_text(size = 9))+
    coord_cartesian(ylim=c(0,35))+
    labs(title='Tooth Length v Vitamin C Dosage by Orange Juice', x='Dose (mg)', y='Odont. Length (micron)')

acidsubset <- ToothGrowth[ToothGrowth$supp == "VC",] ##subset data where supp = VC
acidsubsetplot <- ggplot(acidsubset, aes(x= dose, y=len, fill= dose)) +
    geom_boxplot() +
    theme(plot.title = element_text(size = 10),
         text = element_text(size = 9))+
    coord_cartesian(ylim=c(0,35))+
    labs(title='Tooth Length v Dosage of Vitamin C by Ascorbic Acid', x='Dose (mg)', y='Odont. Length (micron)')
```

```{r , fig.width = 4.1, fig.height = 3.3, echo=FALSE}
orangejplot
acidsubsetplot
```

**2. Provide a basic summary of the data.**

The boxplots appear to show that orange juice delivery provides greater tooth 
length at lower dosages only (0.5 and 1) than ascorbic acid. Increases in dosage
however, results in increases in tooth length for either delivery method.

**3. Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose. **

To compare tooth growth by supplement and dose we will use the `t.test` function 
to determine if the two sets of data are significantly different to each other.

The null hypothesis is that the means of both data sets are equal with the 
alternative hypothesis being that the difference in tooth length is 
statistically significant. The null hypothesis will be rejected if we find a 
p-value of less than 0.05.

Let's first subset the data further.

```{r}
acid05 <- acidsubset[acidsubset$dose==0.5,]
acid10 <- acidsubset[acidsubset$dose==1.0,]
acid20 <- acidsubset[acidsubset$dose==2.0,]
oj05 <- ojsubset[ojsubset$dose==0.5,]
oj10 <- ojsubset[ojsubset$dose==1.0,]
oj20 <- ojsubset[ojsubset$dose==2.0,]
```

First we t-test the effect of supplement type on tooth length while holding 
dosage constant. 

```{r}
# Perform t test on acidsubset vs ojsubset at 0.5mg dose
t.05.acidoj <- t.test(len ~ supp, data=rbind(acid05,oj05), var.equal=FALSE)

#1.0mg dose
t.10.acidoj <- t.test(len ~ supp, data=rbind(acid10,oj10), var.equal=FALSE)

#2.0mg dose
t.20.acidoj <- t.test(len ~ supp, data=rbind(acid20,oj20), var.equal=FALSE)
```

```{r echo=FALSE}

##make display table for data 
a1 <- c("Ascorbic Acid and OJ","Ascorbic Acid and OJ","Ascorbic Acid and OJ")
b1 <- c("0.5mg", "1.0mg", "2.0mg")
c1 <- c(t.05.acidoj$p.value,t.10.acidoj$p.value,t.20.acidoj$p.value)
d1 <- c(t.05.acidoj$conf.int[1],t.05.acidoj$conf.int[1],t.05.acidoj$conf.int[1])
e1 <- c(t.05.acidoj$conf.int[2],t.10.acidoj$conf.int[2],t.20.acidoj$conf.int[2])

resultsummary1 <- data.frame("supplements compared"=a1,
                            "dose level"=b1,
                            "p-value"=c1,
                            "conf. int. (-)"=d1,
                            "conf. int. (+)"=e1, check.names=F)

library(knitr)

kable(resultsummary1, digits = 7, caption = "Summary of Results")

```

Now we'll test the effect of dosages on tooth length:

```{r}
# Perform t test on 0.5mg vs 1.0mg, within each supplement
t.acid.0510 <- t.test(len ~ dose, data=rbind(acid05,acid10), var.equal=TRUE)
t.oj.0510 <- t.test(len ~ dose, data=rbind(oj05,oj10), var.equal=TRUE)

# 1.0mg vs 2.0mg
t.acid.1020 <- t.test(len ~ dose, data=rbind(acid10,acid20), var.equal=TRUE)
t.oj.1020 <- t.test(len ~ dose, data=rbind(oj10,oj20), var.equal=TRUE)

# 0.5mg vs 2.0mg
t.acid.0520 <- t.test(len ~ dose, data=rbind(acid05,acid20), var.equal=TRUE)
t.oj.0520 <- t.test(len ~ dose, data=rbind(oj05,oj20), var.equal=TRUE)
```

```{r echo=FALSE}

##make display table for data 
a <- c("0.5mg and 1.0mg"," 0.5mg and 1.0mg","1.0mg and 2.0mg","1.0mg and 2.0mg", "0.5mg and 2.0mg", "0.5mg and 2.0mg")
b <- c("Acid", "OJ", "Acid", "OJ", "Acid", "OJ")
c <- c(t.acid.0510$p.value,t.oj.0510$p.value,t.acid.1020$p.value,t.oj.1020$p.value,t.acid.0520$p.value,t.oj.0520$p.value)
d <- c(t.acid.0510$conf.int[1],t.oj.0510$conf.int[1],t.acid.1020$conf.int[1],t.oj.1020$conf.int[1],t.acid.0520$conf.int[1],t.oj.0520$conf.int[1])
e <- c(t.acid.0510$conf.int[2],t.oj.0510$conf.int[2],t.acid.1020$conf.int[2],t.oj.1020$conf.int[2],t.acid.0520$conf.int[2],t.oj.0520$conf.int[2])

resultsummary <- data.frame("doses compared"=a,
                    supplement=b,
                    "p-value"=c,
                    "conf. int. (-)"=d,
                    "conf. int. (+)"=e, check.names=F)

kable(resultsummary, digits = 7, caption = "Summary of Results")

```

**4. State your conclusions and the assumptions needed for your conclusions.**

The analysis shows a direct connection between dosage and tooth growth.
p-values for all tests were below threshold and confidence intervals do not 
include zero, indicating that the increase in tooth growth when supplement 
dose is increased is statistically significant.

We see however, that the correlation between supplement type and tooth growth 
diminishes at the highest dose. For 0.5mg and 1.0mg, p-values are below 
threshold and the confidence interval does not include zero, indicating that the
increase in tooth growth that's seen with orange juice vs ascorbic acid is 
statistically significant. At 2.0mg, however, this discrepancy vanishes and our 
null hypothesis at this level is accepted.

---

```{r}

##EXAMPLE SUMMARY TABLE CODE - HIDDEN ABOVE FOR LAYOUT PURPOSES
a1 <- c("Ascorbic Acid and OJ","Ascorbic Acid and OJ","Ascorbic Acid and OJ")
b1 <- c("0.5mg", "1.0mg", "2.0mg")
c1 <- c(t.05.acidoj$p.value,t.10.acidoj$p.value,t.20.acidoj$p.value)
d1 <- c(t.05.acidoj$conf.int[1],t.05.acidoj$conf.int[1],t.05.acidoj$conf.int[1])
e1 <- c(t.05.acidoj$conf.int[2],t.10.acidoj$conf.int[2],t.20.acidoj$conf.int[2])

resultsummary1 <- data.frame("supplements compared"=a1,
                            "dose level"=b1,
                            "p-value"=c1,
                            "conf. int. (-)"=d1,
                            "conf. int. (+)"=e1, check.names=F)
library(knitr)
kable(resultsummary1, digits = 7, caption = "Summary of Results")

```
